Discriminative Sensing Based on Signal Processing

### 2017-10-24 docs/thesis/chapters/appendix/a_mos/a_mos.tex
appendix configuration

\chapter{Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks} 
\label{chapter3}

\section{Introduction}
\label{sec:introduction}

Traditionally, cyber defense methods can be effective against ordinary and conventional types of attacks, yet may fail against innovative malicious techniques \cite{lakhina2005mining}. In order to be able to detect and avoid novel attacks and their variations, it is necessary to develop or improve techniques to achieve efficiency on resource consumption, processing capacity and response time. Moreover, it is crucial to obtain high detection accuracy and capacity to detect variations of malicious patterns. Recently, signal processing schemes have being applied to the detection of malicious traffic in computer networks \cite{Lu2009,Huang2009,Zonglin2009,david2011blind,da2012improved,tenorio2013greatest}, showing advances in network traffic analysis.

Information security may consist of both technical and procedural aspects. The former includes equipment and security systems, while the latter corresponds to security rules and recommendations. Intrusion detection and intrusion prevention systems are security systems used, respectively, to detect (passively) and prevent (proactively) threats to computer systems and computer networks. Such systems can work in the following fashions: signature-based, anomaly-based or hybrid \cite{Huang2009,mudzingwa2012study}. Additionally, anomaly detection techniques can be categorized in classification, statistical, information theory and clustering based, according to \cite{bhuyan2014network,ahmed2016survey,osanaiye2016distributed}.

In the context of anomaly-based schemes, this work proposes a statistical approach based on signal processing techniques for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this work models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). To the best of our knowledge there is no similar model in the literature. The proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy and performance of the proposed framework applied to a experimental scenario and to the DARPA 1998 dataset \citep{osanaiye2016distributed}, which is a well known network traffic dataset. Furthermore, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

The performed experiments show that synflood, fraggle and port scan attacks can be detected accurately and with great detail in an automatic and blind fashion, applying signal processing concepts for traffic modeling and through approaches based on MOS and eigen similarity analysis. The main contributions of the proposed framework are the capability to blindly detect time frames under network attack via MOS and eigen analysis, and the detailed identification of the network attack via eigen similarity analysis.

This paper is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model and the evaluated datasets. Section \ref{sec:prop_getv} describes the proposed framework for blind and automatic detection of flood and probe attacks. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:Complexity} discusses the computational complexity of the proposed framework and evaluates the required processing time for tested scenarios. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

\section{Related Works}
\label{sec:relatedworks}

Several methods have been proposed for the identification and characterization of malicious activity in computer networks. Classical methods typically employ data mining \cite{he2008applying,ghourabi2010data,osanaiye2016distributed} and regular file analysis \cite{raynal2004honeypot} to detect patterns that indicate the presence of specific attacks in network traffic.

Data mining is often used to describe the process of extracting useful information from large databases. Multiple methods of data mining are used in \cite{he2008applying,osanaiye2016distributed} to analyze data flow in a network with the aim of identifying characteristics of malicious traffic in large scale environments. Researchers have applied data mining techniques in log analysis \cite{ghourabi2010data} to improve intrusion detection performance. However, data mining techniques used so far in network analysis require prior collection of large data sets, which is a limitation of several schemes for online analysis \cite{Huang2009}.

Regular file analysis \cite{raynal2004honeypot} consists of traffic analysis for detecting known patterns that indicate the presence of attacks, applying statistical analysis to the study of collected traffic. An essential feature of this method is that it depends on prior knowledge of the details of the attacks to be identified, and also depends on previous log collection for traffic analysis and false positives reduction.

Principal Component Analysis (PCA) is a statistical technique commonly used for dimensionality reduction. It uses an orthogonal transformation to convert a set of correlated variables into a set of linearly uncorrelated variables, where the first principal components have the largest variance. PCA has been used in attack detection \cite{almotairi2009technique}. However, PCA requires human intervention in order to identify abnormalities based on the eigenvalues profiles, if used without complementary techniques.

Callegari \emph{et al} \cite{Zonglin2009} propose a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level and evaluated their proposal through a dataset with synthetic anomalies added in the data. However, Callegari \emph{et al} focus on flood attack detection, not addressing probe attack detection, and their approach relies on visual analysis.

Lee \emph{et al.} \cite{Lee2013} propose OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.}, the framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} only evaluate their proposed scheme for covariance analysis, while we adopt an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for flood and probe attacks, respectively.

Besides being prone to higher errors and false positives, such human intervention makes PCA not useful for real time applications. Therefore, in order to automate the analysis of eigenvalues profile, model order selection (MOS) schemes should be incorporated.

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work requires a training method to produce a prediction model for normal daily traffic and presents limitations on identification of behaviors without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work is not applied to probe and flood attack detection, simultaneously.

The data collected from honeypot systems, such as captured traffic and operating system logs, can be analyzed to obtain information about attack techniques, general trends of threats and exploits. Blind automatic detection of malicious traffic techniques have been developed for honeypots in \cite{david2011blind,da2012improved}. However, traffic on honeypot is simpler than real network traffic, because there are no running legitimate applications, due to the fact that honeypots emulate behavior of a host within a network to deceive and lure attackers \cite{zakaria2012review}. Since honeypots do not generate legitimate traffic, the amount of data captured in honeypots is significantly lower in comparison to a Network Intrusion Detection System (NIDS), which captures and analyzes the largest possible amount of network traffic \cite{david2011blind}. MOS for blind identification of malicious activities in honeypots was proposed by us in \cite{david2011blind}, which evaluated criteria for selecting the model order, through simulations and comparing the order of the resulting model with the true model order.

The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probe and flood attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.

Several approaches for network attack detections uses the KDD 99 \cite{ji2016multi,ahmed2016survey,osanaiye2016distributed,bhuyan2014network} datasets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 dataset are criticized by for the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled datasets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} dataset is the refined version of the KDD 99 dataset that redundant data records are removed, in order to avoid biased classifications. Additionally, some approaches uses simulated \cite{callegari2011novel} scenarios or non-public datasets their evaluations. Since the proposed approach relies on a packet level analysis and the KDD 99 and NSL-KDD datasets adopt a traffic aggregation by connections, we consider the use of a experimental scenario on a real network and the DARPA 1998 dataset, which is the source for the creation of the KDD 99 and NSL-KDD datasets. Note that the proposed approach is not based on learning or classification techniques, which are more susceptible to biased results caused by the issues in the DARPA/KDD datasets.

\section{Data Model}
\label{sec:datamodel}

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $α$, $β$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matriz, and the operator $\sim$ that sorts the elements of a vector in ascending order.

This section presents details of the experimental scenario and the selected cases of the DARPA dataset, along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce flood and probe attacks. Subsection \ref{sec:ModelingData} presents how network traffic can be modeled as signal superposition. Subsection \ref{sec:SynfloodFraggleandPortscan} details the traffic of synflood, fraggle and port scan attacks, and Subsection \ref{sec:DarpaDataset} discusses the use of the DARPA dataset for evaluation of the proposed approach.

\subsection{Analyzed Scenario and Data Collection}
\label{sec:DataCollection}

The environment of the analyzed scenario is composed of two computers and one router with access to the Internet and to an internal network, where the simulation of legitimate traffic, noise, flood and port scan attacks are performed. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:fig1}.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=6cm, width=9cm]{results/figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:fig1}
\end{figure}

Note that the set of network traffic is modeled as legitimate, noise and malicious traffic, where the victim performs legitimate activities, that can be characterized by web access. In many organizations this type of traffic is predominant, since most of corporate services are web-based, such as: web pages, customized web-based systems and cloud services. It is possible to characterize the traffic of a DHCP service as an example of noise associated with the transport layer. For malicious traffic, three types of networks attacks are evaluated: synflood, fraggle and port scan. These attacks are reproduced using well-known security tools, such as Nmap\footnote{http://nmap.org} to port scan, Metasploit\footnote{http://www.metasploit.com} to synflood and Hping\footnote{http://hping.org} to lead the fraggle attack.

A network traffic log is commonly formed by timestamp, protocol, source IP address, source port, destination IP address, destination port and additional information, according to the type of the used transport protocol. The following TCP traffic log is presented in order to exemplify the collected data:
\newline
\newline
\texttt{\justify21:00:34.099289 IP 192.168.1.102.34712 > 200.221.2.45.80: Flags [S], seq 2424058224, win 14600, options [mss 1460, sackOK,TS val 244136 ecr 0,nop,wscale 7], length 0}
\newline

and the following to exemplify UDP traffic log: 
\newline
\newline
\texttt{\justify21:24:42.484858 IP 192.168.1.102.68 > 192.168.1.1.67: BOOTP/DHCP, Request from 00:26:9e:b7:82:be, length 300}
\newline

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e., specific information such as origin IP or day and time of the attack are not considered. Therefore, from the entire log information, we just consider the timestamp (for sequencing), port type and port number.

\subsection{Modeling Data}
\label{sec:ModelingData}

By modeling the dataset as a signal superposition, the network traffic (\textbf{X}) can be characterized as a mixture of three components: legitimate traffic (\textbf{U}), noise (\textbf{N}) and malicious traffic (\textbf{A}), according to the following expression:

\begin{equation}\label{eq:eq01}
\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{N}^{(q)} + \boldsymbol{A}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time grouping of network traffic. The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns. Each row represents a communication port (TCP port or UDP port), and each column represents time bins having a approtiate size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number of times that the port $m$ appears at the $n$-th minute, at the $q$-th time frame.

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's operations. When a user accesses a web page, for example, there is the corresponding TCP/IP traffic to request the page, as well as there is the traffic required to domain name resolution. Figure \ref{fig:fig3} depicts an example of the legitimate traffic obtained during experiments.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:fig3}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:fig4} depicts an example of noise in a network traffic, represented by traffic to ports 67 and 68.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:fig4}
\end{figure}

The traffic coming from a malicious activity, such as a synflood or fraggle attack, is represented by the matrix $\boldsymbol{A}^{(q)}$. For this work we only consider the traffic from port scanning and flood attacks.

We define that if the obtained $\#\boldsymbol{A}^{(q)} ≠ 0$, then there is malicious traffic in the evaluated time frame $q$, on the other hand, if the $\#\boldsymbol{A}^{(q)} = 0$, then there is no malicious traffic. This paper shows how to detect the $\#\boldsymbol{A}^{(q)}$, given only the matrix $\boldsymbol{X}^{(q)}$, in order to identify malicious network traffic.

\subsection{Synflood, Fraggle and Port scan}
\label{sec:SynfloodFraggleandPortscan}

The network attacks evaluated by this work are: synflood, fraggle and port scan. The first two attacks can be qualified as flood or denial of service (DoS) attacks, while the last one can be qualified as probe or port scanning attack. DoS attempts to block access to system or network resources and is implemented by either forcing targets to be unavailable through the exploiting of system vulnerabilities, or consuming resources through large amount of network traffic, characterizing flood attacks. Probe attacks scan the networks to collect information about host, such as IP addresses, ports and services.

With respect to the synflood attacks, the attacker sends a large quantity and concurrent successive SYN requests to a target, in order to consume resources and cause a DoS. Figure \ref{fig:fig5} depicts an example of a synflood attack carried out in a real computer network. In an interval of ten minutes, more than 210,000 packets are sent as a synflood attack. This network traffic behavior can be considered an abnormal behavior of network traffic, especially since it is concentrated in a short period of time and presents similar outstanding traffic during the time under attack.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:fig5}
\end{figure}

With respect to the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo. Figure \ref{fig:fig6} depicts an example of the fraggle attack in a real computer network.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{results/figures/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:fig6}
\end{figure}

More than 6,000,000 malicious packets can be counted in an interval of ten minutes, which can be considered an abnormal network traffic, especially due to the concentrated traffic in a short period of time and due to the similarity of the outstanding traffic.

Port scan is the attempt to establish a connection to TCP and UDP ports to identify what services are running or are in the listening state. There are several available port scanning techniques, including: TCP SYN scan, TCP ACK scan and UDP scan. This work evaluates the use of TCP SYN scan and UDP scan.

In TCP SYN scan, a SYN packet is sent to the destination and two types of responses may occur: SYN/ACK or RST/ACK. In the first case, the destination port is in the listening state, in the second case, the destination port is not listening. At the end of each port scanning, a RST/ACK packet is sent by the system that is performing the port scan. Therefore, a full connection or a complete three-way handshake is never established, which makes the detection of the attack sender more difficult, and requires approaches able to identify probe attacks without connection stablishment. The UDP scan technique sends UDP packets to the destination port, and if it responds with a \emph{ICMP port unreachable} message, then it indicates that the scanned port is closed. On the other hand, if a message is not received, then the port is considered as open.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=5cm, width=9cm]{results/figures/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:fig7}
\end{figure}

Figure \ref{fig:fig7} depicts an example of the port scan attack in a real computer network. Note that the traffic is composed of two packets for each TCP port and one UDP packet to each port. The incoming and outgoing packets analysis, for each port, shows the high correlation and similarity of TCP and UDP traffic during the simulated port scan attack.

\subsection{The DARPA Dataset}
\label{sec:DarpaDataset}

The DARPA 1998 dataset\footnote{https://www.ll.mit.edu/ideval/data/} includes 7 weeks of sniffed traffic saved into raw TCPDUMP packet data, from inside and outside origins, with labeled attacks. The attacks in this dataset
can be grouped into: denial-of-service (DoS); remote to local (R2L), which is characterized by unauthorized access from a
remote machine; user to root (U2R), which is characterized by unauthorized access to local
super-user privileges; and probe attack. Since the proposed approach focus on flood and probe attack, the analysis concentrates on the attacks that present behaviors similar to flood or probe attack. We observe that the most cases of DoS focus on exploit system vulnerabilities instead of on flooding attack. One example is the occurrence of a neptune attack which sends 20 only SYN packets, what is a behavior that differs of the expected flood attack behavior. Therefore, there were selected the cases that simulates several network traffic or numerous connection requests, also known as flooding attack \cite{ahmed2016survey,osanaiye2016distributed}, and the cases that scan ports sending just a few packets. From the simulated probe attacks, we select the cases that rely on TCP or UDP connections.

The data modeling follows the method described by the Subsection \ref{sec:ModelingData}, with time frames of 20 minutes, packet counting aggregation by minute and considering the traffic to the following ports: 20, 21, 22, 23, 25, 79, 80, 88, 107, 109, 110, 113, 115, 143, 161, 389, 443.

\section{Proposed Framework for Detection and Identification of Network Attacks}
\label{sec:prop_getv}

This section describes the proposed technique to detect synflood, fraggle and port scan, according to Figure \ref{fig:fig80}, which represents the overview of the proposed framework for detection and identification of network attacks. In Subsection \ref{sec:prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvelue for each $q$-th time frame. Next, in Subsection \ref{sec:prop_MOSSchemes}, we show how to apply the eiganvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:prop_EigenSimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     \includegraphics[height=11cm, width=9cm]{results/figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:fig80}
\end{figure}

\subsection{Largest Eigenvalue by Time Frames}
\label{sec:prop_LargestEigenvaluebyTimeFrames}

The proposed attack detection algorithm starts by the data pre-processing of a network traffic log containing IP, ports and timestamp of senders and receivers. During this step, the desired information is extracted in order to classify and count packets according to the origin and destination ports, and subsequently this information is grouped by minutes and by time frames.

With the data grouped into $Q$ time frames, the framework considers the time variations of the matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M\times{N}}$, with $q = 1, \ldots, Q$, in order to detect the attack.

According to flood and port scan attacks' behavior, flood attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through the analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:experimentalresults}, which shows that the main components of flood attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

Therefore, to detect flood attacks, it is necessary to calculate the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ of the zero mean samples given by

\begin{equation}\label{eq:eq02}
\boldsymbol{y}_{m}^{(q)} = \boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}.
\end{equation}

The set of obtained vectors $\boldsymbol{y}_{m}^{(q)}$ composes the zero mean matrix $\boldsymbol{Y}^{(q)}$, then the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ can be calculated as follows

\begin{equation}\label{eq:eq03}
\boldsymbol{\hat{R}}_{yy}^{(q)} = \frac{1}{N}\boldsymbol{Y}^{(q)}\boldsymbol{Y}^{(q)^{\rm T}}.
\end{equation}

For the detection of the port scan attack, the main components are not dominated by the variables with large variance. Moreover, the portscan traffic presents a highly correlated network traffic. In order to exploit such structure, we compute the sample covariance $\boldsymbol{\hat{R}}_{zz}^{(q)}$ whose variables have zero mean and unitary standard deviation as follows

\begin{equation}\label{eq:eq04}
\boldsymbol{z}_{m}^{(q)} = \frac{\boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}}{\boldsymbol{\sigma}_{m}^{(q)}}.
\end{equation}

The set of vectors $\boldsymbol{z}_{m}^{(q)}$ composes the matrix $\boldsymbol{Z}^{(q)}$, then the sample covariance matrix $\boldsymbol{\hat{R}}_{zz}^{(q)}$ can be calculated via

\begin{equation}\label{eq:eq05}
\boldsymbol{\hat{R}}_{zz}^{(q)} = \frac{1}{N}\boldsymbol{Z}^{(q)}\boldsymbol{Z}^{(q)^{\rm T}}.
\end{equation}

Once the $\boldsymbol{\hat{R}}_{yy}^{(q)}$ and $\boldsymbol{\hat{R}}_{zz}^{(q)}$ have been obtained for flood and port scan attack detection, respectively, and since the next steps are the same for both sample covariance matrices, we refer to $\boldsymbol{\hat{R}}_{yy}$ and $\boldsymbol{\hat{R}}_{zz}$ as a matrix $\boldsymbol{C}$. Therefore, the following step of the algorithm is the eigenvalue decomposition (EVD), calculated according to (\ref{eq:eq06}), in order to obtain the vector of eigenvalues $\boldsymbol{e}^{(q)}$ associated with each matrix, according to (\ref{eq:eq060}).

\begin{equation}\label{eq:eq06}
\boldsymbol{C}^{(q)} = \boldsymbol{V}^{(q)}\boldsymbol{\Lambda}^{(q)}\boldsymbol{V}^{(q)^{\rm T}},
\end{equation}

\begin{equation}\label{eq:eq060}
\boldsymbol{e}^{(q)} = \rm diag(\boldsymbol{\Lambda}^{(q)}),
\end{equation}

where the operator diag$(\cdot)$ extracts the main diagonal of a matrix.

The eigenvalues should be sorted in descending order, i.e., $\lambda_{1}^{(q)} > \lambda_{2}^{(q)} > \lambda_{3}^{(q)} > ... > \lambda_{m}^{(q)}$. Therefore, the largest eigenvalue of the $q$-th time frame evaluated for the attack detect is given by $\lambda_{1}^{(q)}$.

The concatenation of the eigenvalues vector $\boldsymbol{e}^{(q)}$ for $q = 1, \ldots, Q$ is represented by

\begin{equation}\label{eq:eq07}
\boldsymbol{E} =
\begin{bmatrix}
  \lambda_1^{(1)} & \lambda_1^{(2)} & \lambda_1^{(3)} & \cdots & \lambda_1^{(Q)} \\
  \lambda_2^{(1)} & \lambda_2^{(2)} & \lambda_2^{(3)} & \cdots & \lambda_2^{(Q)} \\
  \lambda_3^{(1)} & \lambda_3^{(2)} & \lambda_3^{(3)} & \cdots & \lambda_3^{(Q)} \\
  \vdots & \vdots & \ddots & \vdots  \\
  \lambda_m^{(1)} & \lambda_m^{(2)} & \lambda_m^{(3)} & \cdots & \lambda_m^{(Q)} \\
\end{bmatrix}.
\end{equation}

Note that since $\lambda_1^{(q)} > \lambda_2^{(q)} > \lambda_3^{(q)} > \cdots > \lambda_{m-1}^{(q)} > \lambda_m^{(q)}$, then the first line of the matrix $\boldsymbol{E}$ contains the largest eigenvalues of each $q$-th time frame, which is the Greatest Eigenvalue Time
Vector (GETV) \cite{tenorio2013greatest}, denoted as

\begin{equation}\label{eq:eq08}
\boldsymbol{e}_{\rm max} = \boldsymbol{E}\{:,1\} = [ \lambda_1^{(1)}, \lambda_1^{(2)} ... \lambda_1^{(Q)}]
\end{equation}

\subsection{MOS Schemes}
\label{sec:prop_MOSSchemes}

Traditionally the MOS schemes are applied for the eigenvalues of the vector $\boldsymbol{e}^{(q)}$. However, the goal here is to detect the variations of the eigenvalues for different values of $q$. Therefore, instead of using a certain $q$, the proposed approach applies MOS schemes for a vector of the largest eigenvalues of each $q$-th time frame, in order to identify variations and estimate the model order $\hat{d}$, which is the estimated number of time frames under attack. Therefore, $\boldsymbol{e}_{\rm max}$ is sorted in descending order, producing $\sim\boldsymbol{e}_{\rm max}$, that is used as input parameter for MOS schemes, according to $\hat{d} = \rm{MOS}(\sim\boldsymbol{e}_{\rm max})$. Note that some MOS schemes may also require the number of minutes that compose a time frame, as $\hat{d} = \rm{MOS}(\boldsymbol{e}_{\rm max},\emph{Q})$. For more information about MOS, we refer to \ref{sec:mos}.

In our previous work \cite{tenorio2013greatest}, the accuracy of AIC, MDL, EDC, RADOI, EFT and SURE schemes are evaluated for synflood and port scan attack detection, showing that EDC and EFT are effective for detecting this kind of attacks. The present work extends that evaluation to also analyze the effectiveness of the listed MOS schemes for fraggle attack detection, as shown in Section \ref{sec:experimentalresults}.

\subsection{Eigenvalue Analysis}
\label{sec:prop_EigenvalueAnalysis}

After applying the MOS schemes to the vector $\sim\boldsymbol{e}_{\rm max}$, we obtain the estimate of the $\#\boldsymbol{A}$. For instance, in the case of fraggle, synflood and portscan, if $\hat{d} = 1$, , then $\#\boldsymbol{A} = 1$, which means that during the during the $Q$ time frames one attack is present. However, if $\hat{d} = 0$, then $\#\boldsymbol{A} = 0$, and this means that none of these attacks are present. Note that $\hat{d}$ can be greater than 1, indicating the presence of more than one attack.

In Subsection \ref{sec:prop_MOSSchemes}, we obtained only if $\hat{d} = 1$ or $\hat{d} = 0$. However, if $\hat{d} = 1$, the MOS schemes do not provide any information about the $q$-th time frame under attack. The identification of the $q$-th time frame under attack can be carried out through a eigenvalues analysis.

The largest eigenvalue analysis for estimating the $q$-th time frames that are under attack can be expressed according to Algorithm \ref{alg:alg01}, where $\rm{\boldsymbol{\hat{q}}}_{\rm max} \in \mathbb{R}^{\hat{d}}$ denotes a vector of the $q$-th time frames under attack, which is the $q$-th indexes corresponding to the $\hat{d}$ largest eigenvalues of $\boldsymbol{e}_{\rm max}$. Algorithm \ref{alg:alg01} initially identifies the largest value of $\boldsymbol{e}_{\rm max}$, according to Line 2 of Algorithm \ref{alg:alg01}, and its correspondent index, according to Lines 4 and 5 of Algorithm \ref{alg:alg01}. Subsequently, the largest value is removed of $\boldsymbol{e}_{\rm max}$, according to Line 8 of Algorithm \ref{alg:alg01}, and a new iteration is performed until $\boldsymbol{e}_{\rm max} = \{\}$.

\begin{algorithm}[h!]
	\caption{Detection of Time Frames Under Attack}
  	\label{alg:alg01}
	\begin{algorithmic}[1]
		\show\LOOP
	    \ALOOP {$f = 1$ until $f == \hat{d}$} 
	    	    \STATE $q_{\rm value} = \argmax\limits_{\lambda}  \hspace{1 mm} \boldsymbol{e}_{\rm max}$
    	    	    \ALOOP {$i = 1$ until $i == Q$} 
				\IF {$\boldsymbol{e}_{\rm max}^{(i)} == q_{\rm value}$}
				    \STATE $\boldsymbol{\hat{q}}_{\rm max}^{(f)} = i$
				\ENDIF
        		\ENDALOOP
	    		\STATE $\boldsymbol{e}_{\rm max} \rightarrow \boldsymbol{\hat{q}}_{\rm max}^{(f)}$
    		\ENDALOOP
	\end{algorithmic}
\end{algorithm}

After the estimation of the $\rm{\boldsymbol{\hat{q}}}_{\rm max}$ time frames under attack, it is necessary to obtain more details of the detected attacks, such as the $n$-th minutes when the attacks happened and the $m$-th network ports that were attacked. To deal with this problem, the adoption of a similarity analysis between legitimate traffic and the traffic of time frames estimated as under attack is evaluated, analysing the effectiveness of cosine similarity to highlight abnormalies inserted by network traffic attacks.

\subsection{Eigen Similarity Analysis}
\label{sec:prop_EigenSimilarityAnalysis}

Cosine similarity calculates the cosine of the angle between two vectors, which represents the similarity of values between the selected vectors. Therefore, cosine similarity can be used to evaluate the variation of the most significant eigenvectors of $\boldsymbol{V}^{(q)}$ against the the most significant eigenvectors of time frame detected as under attack, to analyze similarity changes into the most significant eigenvectors caused by the insertion of anomalous traffic \cite{Lee2013}.

This subsection describes the proposed eigen similarity analysis for detailed attack identification, in complement to the attack estimation carried out through MOS schemes and eigenvalue analysis. In Subsection \ref{sec:prop_TimeSimilarityAnalysis} we present the eigen similarity analysis for identification of time under attack. Next, in Subsection \ref{sec:prop_PortSimilarityAnalysis}, we show how to apply the eigen similarity analysis in order to identify network ports under attack.

\subsubsection{Time Similarity Analysis}
\label{sec:prop_TimeSimilarityAnalysis}

For eigen similarity analyis, we evaluate the cosine similarity in order to identify lacks of similarity between legitimate and malicious traffic, as follows

\begin{equation}\label{eq:eq11}
s_n = \frac{\abs{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(n)}}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(n)}}},
\end{equation}
where $s_n$ denotes the absolute similarity degree of the $n$-th minute, $\boldsymbol{v}^{(q)}$ is the most significant eigenvectors of a selected set of minutes without network attack, and $\boldsymbol{v}_{(n)}$ is the most significant eigenvectors obtained after append the target $n$-th minute of traffic to be performed the flood and port scan attack identification.

The most significant eigenvector $\boldsymbol{v}^{(q)}$, of a time frame $q$ without attack, can be derivated from (\ref{eq:eq06}) and selected according to the eigenvector of the largest eigenvalue $\lambda_1^{(q)}$, which is the principal component of the evaluated matrix. The same calculation shall be performed in order to obtain the target eigenvectors $\boldsymbol{v}_{(n)}$, calculated from a time frame without attack plus minutes of a time frame estimated as under attack, to evaluate the occurrence of network attacks.

The reference eigenvectors $\boldsymbol{v}^{(q)}$ is calculated from the traffic whithout attack, from a time frame $q$ composed of $Q$ minutes of legitimate network traffic. For the detailed attack identification, each $\boldsymbol{x}^{(\hat{q})}_{(n)}$ vector of each $n$-th minutes of the estimated $\rm{\boldsymbol{\hat{q}}}_{\rm max}$ time frames shall be individually appended into $\boldsymbol{X}^{(q)}$, as represented by

\begin{equation}\label{eq:eq12}
\boldsymbol{X}_{n} = \{\boldsymbol{X}^{(q)} | \boldsymbol{x}^{(\hat{q})}_{(n)}\}.
\end{equation}

The resultant $\boldsymbol{X}_{(n)}$ is necessary to obtain $\boldsymbol{v}_{(n)}$, through (\ref{eq:eq06}), for calculating the similarity degree $s_n$, ranging from 0 to 1, for each $n$-th minute. The $s_n$ denotes the absolute similarity degree of the $n$-th minute in comparison to a well-known traffic without attack, detected through MOS schemes and eigenvalue analysis.

The incremental approach for similarity analysis is based on the incremental appending of network traffic into $\boldsymbol{X}^{(q)}$, where the first evaluation is based on (\ref{eq:eq12}) and the subsequent evaluations is based on (\ref{eq:eq13}), incrementally appending each $n$-th minute until $n=N$.

>>>>

\chapter{Model Order Selection (MOS)}
\label{sec:a_mos}

The model order selection is a key point in many digital signal processing applications, including radar, sonar, communications, channel modeling, medical imaging, among others. MOS allows analysis of reduced data set, through separating noise components of the main components, for example. Moreover, the model order is crucial for many parameter estimation techniques \cite{da2009comparison}, since the amount of parameters to be estimated depends on the model order.
